---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# mantar - Missingness Alleviation for NeTwork Analysis in R

<!-- badges: start --> 
<!-- badges: end -->

`mantar` provides users with several methods for handling missing data in the context of network analysis. Currently, these methods are specifically implemented for network estimation via neighborhood selection using the Bayesian Information Criterion (BIC).

## Installation

You can install the development version of mantar from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("kai-nehler/mantar")
```

After installation the easiest way to get an overview of functions and capabilities is to use `?mantar` to open the package help-file. You could also read the rest of this README for an introduction and some examples.

## Features

As already described, the package currently focuses on network estimation using **neighborhood selection** with **information criteria** for model selection in node-wise regressions. This functionality is available for both complete and incomplete data.

For datasets with missing values, two modern missing approaches are implemented:

- **Two-step Expectation-Maximization (EM):** A fast method that estimates the correlation matrix via an EM algorithm using the `lavaan` package. It performs well when the sample size is very large relative to the amount of missingness and the complexity of the network.
- **Stacked Multiple Imputation (MI):** A more robust approach across a wider range of sample sizes. Multiple imputation is performed using predictive mean matching (PMM) with the `mice` package. The imputed data sets are stacked into a single data set, and a correlation matrix is estimated from this combined data.

Both methods produce a correlation matrix that is then used to estimate the network via node-wise regressions. It is also possible to compute the correlation matrix using pairwise or listwise deletion. However, these methods are generally not recommended, except in specific cases, such as when data are missing completely at random and the proportion of missingness is very small.

In addition to network estimation, the package also supports **stepwise regression search** based on information criteria for a **single dependent variable**. This regression search is available for both complete and incomplete data and relies on the same two-step EM or stacked MI procedures to handle missing values as the network analysis.  While both methods to handle missingness are expected to perform well in this context, no specific simulation study has been conducted to compare their effectiveness for single regression modeling, and thus their relative strengths remain an open question.

## Example

The package includes two dummy datasets that resemble a typical psychological dataset, where the number of observations is considerably larger than the number of variables. Although the variables have descriptive names, these are included solely to make the examples more engaging - the data themselves are **fully synthetic**.

- `mantar_dummy_full`: Fully observed data (no missing values)
- `mantar_dummy_mis`: Data with missing values

These data sets are intended for examples and testing only.

```{r example}
library(mantar)

# Load example data
data(mantar_dummy_full)
data(mantar_dummy_mis)

# Preview the first few rows
head(mantar_dummy_full)
head(mantar_dummy_mis)
```

The main function for estimating a network is `neighborhood_net()`. In the case of fully observed data, the function takes the dataset as input and estimates a network structure using **neighborhood selection** guided by **information criteria**. With default arguments, only the dataset needs to be provided.

#### Information Criteria

The `k` argument controls the penalty used in model selection for node-wise regressions. It reflects the penalty per parameter (i.e., number of predictors + 1):

- `k = "log(n)"` (default): corresponds to the **Bayesian Information Criterion (BIC)**
- `k = "2"`: corresponds to the **Akaike Information Criterion (AIC)**

#### Estimation of Partial Correlation 

The `pcor_merge_rule` argument determines how partial correlations are estimated based on the regression results between two nodes:

- `"and"` (default): a partial correlation is estimated **only if both** regression weights (from node A to B and from B to A) are non-zero.
- `"or"`: a partial correlation is estimated if **at least one** of the two regression weights is non-zero.

Although both options are available, current simulation evidence suggests that the `"and"` rule yields more accurate partial correlation estimates than the `"or"` rule. Therefore, changing this default is **not recommended** unless you have a specific reason.

#### Type of Correlation

The `cor_method` argument specifies the type of correlation to be estimated. Currently, only two types are supported in this context, as these were evaluated in the given setting. 

- `"pearson"`: computes the **Pearson correlation**.
- `"polychoric"`: computes the **polychoric correlation**. 

The default for this argument is `"adapted"`, which automatically selects an appropriate correlation type based on the data. If all variables are continuous, `"pearson"` is used. If at least one variable is ordinal, the procedure considers the number of variables, the number of categories, and the sample size to decide whether to apply `"pearson"` or `"polychoric"`. If you want to force a specific type of correlation, you can set `cor_method` to either `"pearson"` or `"polychoric"`.

### Example of Network Estimation without Missing Data

```{r example_network}
# Estimate network from full data set using BIC, the and rule as well as an adapted correlation type
result <- neighborhood_net(data = mantar_dummy_full, 
                           k = "log(n)", 
                           pcor_merge_rule = "and",
                           cor_method = "adapted")
# View estimated partial correlations
result

# Create and view a summary of the network estimation
sum_result <- summary(result)
sum_result
```


In the case of missing data, the `neighborhood_net()` function offers several additional arguments that control how sample size and missingness are handled.

#### Calculation of Sample Size

The `n_calc` argument specifies how the sample size is calculated for each node-wise regression. This affects the penalty term used in model selection.

The available options are:

- `"individual"` *(default)*: Uses the number of non-missing observations for each individual variable. This is the recommended approach.
- `"average"`: Uses the average number of non-missing observations across all variables.
- `"max"`: Uses the maximum number of non-missing observations across all variables.
- `"total"`: Uses the total number of observations in the dataset (i.e., the number of rows).

#### Handling Missing Data

The `missing_handling` argument specifies how the correlation matrix is estimated when the input data contains missing values. Two approaches are supported:

- `"two-step-em"`: Applies a standard **Expectation-Maximization (EM)** algorithm to estimate the covariance matrix.
- `"stacked-mi"`: Applies **multiple imputation** to create several completed datasets, which are then stacked into a single dataset. A correlation matrix is computed from this stacked data. 

As described previously, deletion techniques (listwise and pairwise) are also available, but their use is not recommended. When `"two-step-em"` is selected, the correlation matrix is always based on Pearson correlations. 

If `"stacked-mi"` is used, the `nimp` argument controls the number of imputations (default: `20`), while `imp_method` specifies the imputation method (default: `"pmm"` for predictive mean matching).  


### Example of Network Estimation with Missing Data

```{r example_network_mis}
# Estimate network for data set with missing values
result_mis <- neighborhood_net(data = mantar_dummy_mis, 
                                n_calc = "individual", 
                                missing_handling = "two-step-em",
                                pcor_merge_rule = "and",
                                cor_method = "adapted")
# View estimated partial correlations
result_mis

# Create and view a summary of the network estimation
sum_result_mis <- summary(result_mis)
sum_result_mis
```
